Parallel Adaptive Importance Sampling

### 2016-12-04 PAIS.tex
added section on AIS

>>>>

\subsection{Adaptive Importance Sampling}

Bayesian philosophy has become wide\-spread in statistical inference
largely due to the development of efficient methods, such as MCMC and
particle filtering, for calculating previously intractable integrals.
A related class of algorithms, adaptive importance sampling (AIS)
\cite{liu2008monte} reviewed in\cite{bugallo2015adaptive}, had
received less attention until their practical applicability was
demonstrated in the
mid-2000s\cite{celeux2006iterated,cappepopulation,isard1998condensation,bink2008bayesian}.
AIS methods produce a sequence of approximating distributions,
constructed from standard distributions, from which samples can be
easily drawn. At each iteration the samples are weighted, often using
standard importance sampling methods. The weighted samples are used to
train the adapting sequence of distributions so that samples are drawn
more efficiently as the iterations progress. The weighted samples form
a sample from the posterior distribution under some mild
conditions~\cite{robert2013monte,martino2015adaptive}.

The AIS algorithm is Markovian, conditioned on the method of adaption,
but differs from MCMC methods in that the sequence does not have the
posterior as its limiting distribution. Convergence is ensured by the
weighting of samples, which does not rely on algorithm asymptotics.
This means that after any iteration the algorithm can be stopped and
the sample obtained is a sample from the posterior distribution, with
no consideration for burn-in times~\cite{douc2007convergence}.

A popular AIS algorithm is the population monte carlo (PMC) sampling
method~\cite{cappe2012population}. This algorithm uses a population of
samples to build a mixture or kernel density estimate (KDE) of the
posterior distribution. This KDE is optimal in the sense that it
minimises the Kullback-Leibler divergence between the mixture and the
posterior distribution~\cite{douc2007convergence}. The efficiency of
this optimisation is restricted by the component kernel(s) chosen, and
the quality of the current sample from the posterior.

The PMC algorithm is often implemented as a $D$-kernel
method~\cite{douc2007convergence}. In this method one choses $D$
different measures, and a number of samples $N$ such that the total
population size $N = MD$. At each iteration and for each mixture
component, one of these $D$ kernels is selected via some optimisation
criterion and $M$ samples are produced from this kernel. The $M$
samples from each kernel attempt to stabilise the resampling scheme so
that information about a particular region is not lost.

In practice, the PMC algorithm can suffer from spiky posteriors due to
poor approximation properties in some parts of the posterior. The
adaptive multiple importance sampling algorithm
(AMIS)~\cite{cornuet2012adaptive} aims to reduce the impact of these
poor approximations. AMIS considers the entire sample at time $n$ to
be a draw from all approximations up to time $n$. This means that the
importance weights for the whole sample must be recalculated at each
iteration. These weights are known as deterministic mixture
weights~\cite{owen2000safe,veach1995optimally}, where the denominator
is the sum of all kernels, $q_t(\cdot), 0\leq t\leq n$,
\[
	w_n = \pi(Y_n) \Big/ \sum\limits_{t=0}^n \! q_t(Y_n).
\]
The advantage of this system is that at a time $n_1$ a region might be
badly approximated, it might be better approximated at a later time
$n_2$, so the weight for the sample $Y_{n_1}$ will be more
representative of the density in that region, if it is considered as a
draw from a mixture of samples up to time $n_2$. These weights are
however biased, consistency was proved in~\cite{marin2014consistency}.
AMIS has been successfully implemented for problems in population
genetics~\cite{siren2011reconstructing}.

A combination of AMIS and PMC results in a stable PMC algorithm known
as adaptive population importance sampling
(APIS)~\cite{martino2015adaptive}. This algorithm utilises the
deterministic mixture weights to stabilise the algorithm within a
single population, then `restarts' the kernel approximation in the
next iteration. This memoryless property makes the algorithm much more
flexible to adaptation. A combination with the $D$-kernel PMC method
allows for adaptive learning of kernel, kernel mean and kernel
covariances.
